Denotational Semantics of Gradual Typing using Synthetic Guarded Domain Theory

### 2024-07-09 paper-new/concrete-model.tex
work on relational model

In this section, we complete the construction of a relational model of gradual
typing. To manage the complexity, we carry out the construction in phases,
beginning with a relational model consisting of partially-ordered sets with a
notion of bisimilarity and then adding requirements for a structured set of
delays on each type as well as a notion of representability for relations that
models the cast rules.

@@

\subsubsection{Guarded Lift Monad and Free-Forgetful Adjunction}\label{sec:guarded-lift-monad}

@@

\subsection{Phase 2: Perturbations and Quasi-Representable Relations}

@@

We must now impose an additional condition on predomain relations $c$ and error
domain relations $d$ that specify how they interact with these perturbations. In
particular, we need to be able to ``push'' and ``pull'' perturbations along
relations $c$ and $d$. The intuition for this requirement comes from the
construction of the square corresponding to the original UpL rule from the
square for the simplified version as was shown in Section
\ref{sec:modeling-term-precision}. Recall that we horizontally composed the
simplified UpL for $c$ square with the identity square for $c'$ on the right.
Now that the square for UpL involves a delay, this construction doesn't work
unless we can ``push'' the delay on $A_2$ to a delay on $A_3$. That is, given a
relation $c : A \rel A'$ and a perturbation $m_A \in M_A$, we need to be able to
turn it into a perturbation $m_{A'} \in M_{A'}$ such that the resulting
endomorphisms obtained by the respective homomorphisms $i_A$ and $i_{A'}$ form a
square. This is made formal by the following definition:
%
\begin{definition}
    Let $(A, M_A, i_A)$ and $(A', M_{A'}, and i_{A'})$ be value objects, and let
    $c : A \rel A'$ be a predomain relation. A \emph{push-pull structure} for
    $c$ consists of a homomorphism $\push : M_A \to M_A'$ and $\pull : M_A' \to
    M_A$ such that for all $m_A \in M_A$ there is a square $i_A(m_A) \ltsq{c}{c}
    i_{A'}(\push\, m_A)$ and for all $m_{A'} \in M_{A'}$ there is a square
    $i_A(\pull\, m_{A'}) \ltsq{c}{c} i_{A'}(m_{A'})$.
\end{definition}
%
We make the analogous definition for computation objects.

With the notion of perturbation defined, we can now discuss the final aspect of
the model, namely the representability of relations.
%
\begin{definition}[quasi-left-representable relations]
Let $(A, M_A, i_A)$ and $(A', M_{A'}, i_{A'})$ be value objects, and let $c : A
\rel A'$ be a predomain relation. We say that $c$ is
\emph{quasi-left-representable} by a predomain morphism $e : A \to A'$ if there
are perturbations $\delle_c \in M_A$ and $\delre_c \in M_{A'}$ such that the
following two squares exist: (UpL) $e \ltsq{c}{r(A')} i_{A'}(\delre_c)$, and
(UpR) $i_A(\delle_c) \ltsq{r(A)}{c} e$.
\end{definition}
%
Observe that this generalizes the previous notion of representability, since
under that definition the perturbations were always the identity.
%
We make the analogous definition of quasi-left-representability for error domain
relations $d$.
%
Likewise, we define quasi-right-representability as follows:
%
\begin{definition}
Let $(B, M_B, i_B)$ and $(B', M_{B'}, i_{B'})$ be computation objects, and let
$d : B \rel B'$ be an error domain relation. We say that $d$ is
\emph{quasi-right-representable} by an error domain morphism $p : B' \to B$ if
there are perturbations $\dellp_d \in M_B$ and $\delrp_d \in M_{B'}$ such that
the following two squares exist: 
(DnL) $p \ltsq{r(B')}{d} i_{B'}(\delrp_d)$ and
(DnR) $i_B(\dellp_d) \ltsq{d}{r(B)} p$.
\end{definition}
%
We make the analogous definition of quasi-right-representability for predomain
relations $c$.

@@

We can now give the final definition of value and computation relations for our
model:
%
\begin{definition}[value relations]
A \textbf{value relation} between value objects $(A, M_A, i_A)$ and $(A', M_{A'},
i_{A'})$ is a predomain relation $c : A \rel A'$ that has a push-pull structure
is quasi-left-representable by a morphism $e_c : A \to A'$, and is such that
$Fc$ is quasi-right-representable by an error domain morphism $p_c : FA' \to
FA$.
\end{definition}

\begin{definition}[computation relations]
A \textbf{computation relation} between computation objects $(B, M_B, i_B)$ and $(B',
M_{B'}, i_{B'})$ is an error domain relation $d$ that has a push-pull structure,
is quasi-right-representable by a morphism $p_d : B' \to B$, and is such that
$Ud$ is quasi-left-representable by a morphism $e_d : UB \to UB'$
\end{definition}

Because we have augmented the definitions of object and relation for our model,
we need to specify the actions of the functors $F$, $U$, $\times$, and $\arr$ on
these objects and relations. For instance, the functor $F$ now acts on
\emph{value objects} $(A, M_A, i_A)$, which means we need to specify the monoid
$M_{FA}$ and interpretation $i_{FA}$ corresponding to the error domain $FA$. We
leave these constructions to the appendix, as they are technical and not crucial
to understanding our main result.

\subsection{The Dynamic Type}
Lastly, we describe the denotation of the dynamic type. The predomain
representing the dynamic type is defined using guarded recursion as the solution
to the equation
% \footnote{In this section, we write $F$ instead of $\li$ so that the notation follows that of the abstract model section.}
%
\[ D \cong \mathbb{N}\, + (D \times D)\, + \laterhs U(D \arr FD). \]
%
% Note that the operators in the above equation are all combinators for predomains, so
% this also defines the ordering and the bisimilarity relation for $D$.

@@

Now we define the perturbations for $D$. Recall that for each value type $A$ we
associate a monoid $P_A$ of perturbations and a homomorphism into the monoid of
endomorphisms bisimilar to the identity, and likewise for computation types. We
define the perturbations for $D$ via least-fixpoint in the category of monoids
as
%
\( P_D \cong (P_{D \times D}) \times P_{U(D \to FD)}. \)
%
Unfolding the definitions, this is the same as
%
\( P_D \cong (P_D \times P_D) \times (\mathbb{N} \times P_D^{op} \times \mathbb{N} \times P_D). \)
%
We now explain how to interpret these perturbations as endomorphisms.
We define $\ptb_D : P_D \to \{ f : D \to D \mid f \bisim \id \}$ below,
% via the universal property of the coproduct of monoids, giving a case for each of the generators.
% In the below, note the use of the functorial action of $\arr$ on morphisms.
%
% \[ P_D \cong (P_D \times P_D) \times (P_D \times (\mathbb{N} \times P_D)), \]
%
\begin{align*}
 \ptb_D(p_{\text{times}}, p_{\text{fun}}) &= \lambda d.\text{case $d$ of}  \\
 &\alt \tnat(m) \mapsto \tnat(m) \\
    &\alt \ttimes(d_1, d_2) \mapsto {\ttimes(\ptb_{D \times D}(p_\text{times})(d_1, d_2))} \\
    &\alt \tfun(\tilde{f}) \mapsto {\tfun(\lambda t. \ptb_{U(D \to FD)}()(\tilde{f}_t))}
\end{align*}
    % \item $\ptb_D(1)$
    % \item $\ptb_D(\delta^K_D)$ is defined similarly to the previous but has
    % \[ \id \arr i^K(\delta^K_D) \] instead.
    % \item $\ptb_D(\delta^K_D) = \lambda d.\text{case $d$ of}$
    %   \begin{align*} 
    %     &\alt \ttimes(d_1, d_2) \mapsto {\ttimes(i^K(\delta^K_D)(d_1), d_2)} \\
    %     &\alt d' \to d'
    %   \end{align*}
    % \item $\ptb_D(\delta^K_D)$ is defined similarly to the previous but has 
    % \[ (d_1, \ptb_D(\delta^K_D)(d_2)) \] instead.
One can verify that this defines a homomorphism from $P_D \to \{ f : D \to D : f \bisim \id \}$.
We claim that the three relations $\inat$, $\itimes$, and $\iarr$
%and their lifted versions 
satisfy the push-pull property.
As an illustrative case, we establish the push-pull property for the relation $\iarr$.
We define $\pull_{\iarr} : P_D \to P_{U(D \arr FD)}$ by \( \pull_{\iarr}(p_{\text{times}}, p_{\text{fun}}) = p_{\text{fun}}, \)
% (recall that $P_{U(D \arr FD)} = \mathbb{N} \times P_D^{op} \times \mathbb{N} \times P_D$).
%
i.e., we simply forget the other perturbation.
%
We define $\push_{\iarr} : P_{U(D \arr FD)} \to P_D$ by \( \push_{\iarr}(p_{\text{fun}}) = (\id, p_{\text{fun}}) \).
%
Showing that the relevant squares commute is straightforward.
%
% To do so, consider an arbitrary perturbation $p$ on $D$. 
% Let $(f, d) \in \iarr$. This means that $d$ must be of the form $\tfun{\tilde{f}}$. 
% When we interpret the perturbation on $D$, obtaining and endomorphism to which we then apply $d$,
% we will be perform the action that was performed by the perturbation on the other side and thus
% we will be done by our assumption that $f$ is related to $d$.

@@

Now that we have defined an intensional model with an interpretation for the dynamic type, we can apply
the abstract constructions introduced in Section \ref{sec:extensional-model-construction}.
Doing so, we obtain an extensional model of gradual typing, where the squares are given by the
``bisimilarity closure'' of the intensional error ordering.

@@

In this section, we prove an adequacy result for the concrete extensional model of GTT we obtained in the previous section.
applying the abstract constructions introduced in Section
\ref{sec:extensional-model-construction} to the concrete model built in the previous section.
%\ref{sec:concrete-model}.

First we establish some notation. Fix a morphism $f : 1 \to \li \mathbb{N} \cong \li \mathbb{N}$.
We write $f \da n$ to mean that there exists $m$ such that $f = \delta^m(\eta n)$
and $f \da \mho$ to mean that there exists $m$ such that $f = \delta^m(\mho)$.

Recall that $\ltls$ denotes the relation on value morphisms defined as the bisimilarity-closure
of the intensional error-ordering on morphisms.
That is, we have $f \ltls g$ iff there exists $f'$ and $g'$ with
%
\[ f \bisim_{\li \mathbb{N}} f' \le_{\li \mathbb{N}} g' \bisim_{\li \mathbb{N}} g. \]
%
Here $\le_{\li \mathbb{N}}$ is the lock-step error ordering, and
$\bisim_{\li \mathbb{N}}$ is weak bisimilarity.
First observe that in this ordering, the semantics of error is not equivalent to
the semantics of the diverging term.
The main result we would like to show is as follows:
\begin{lemma}
If $f \ltls g : \li \mathbb{N}$, then:
\begin{itemize}
  \item If $f \da n$ then $g \da n$.
  \item If $g \da \mho$ then $f \da \mho$.
  \item If $g \da n$ then $f \da n$ or $f \da \mho$.
\end{itemize}
\end{lemma}
%
Unfortunately, this is actually not provable!
Roughly speaking, the issue is that this is a ``global'' result, and it is not possible
to prove such results inside of the guarded setting. 
In particular, if we tried to prove the above result in the guarded
setting, we would run into a problem where we would have a natural number
``stuck'' under a $\later$, with no way to get out the underlying number.
%
Thus, to prove our adequacy result, we need to leave the guarded setting and pass back
to the more familiar, set-theoretic world with no internal notion of step-indexing.
We can do this using a process known as \emph{clock quantification}.
Recall that all of the constructions we have made in SGDT take place in the context of a clock $k$.
All of our uses of the later modality and guarded recursion have taken place with respect to this clock.
For example, recall the definition of the lift monad by guarded recursion.
% We define the lift monad $\li^k X$ as the guarded fixpoint of $\lambda \tilde{T}. X + 1 + \later^k_t (\tilde{T}_t)$.
We can view this definition as being parameterized by a clock $k$: $\li^k : \type \to \type$.
Then for $X$ satisfying a certain technical requirement known as \emph{clock-irrelevance},
\footnote{A type $X$ is clock-irrelevant if there is an isomorphism $\forall k.X \cong X$.}
we can define the ``global lift'' monad as $\li^{gl} X := \forall k. \li^k X$.

>>>>

In this section, we complete the construction of a relational model of gradual
typing. To manage the complexity, we carry out the construction in phases,
beginning with a relational model consisting of partially-ordered sets with a
notion of bisimilarity and then adding requirements for a structured set of
delays on each type as well as a notion of representability for relations that
models the cast rules in the intensional setting.

@@

\subsubsection{The Free Error Domain}\label{sec:free-error-domain}

@@

\subsection{Phase Two: Perturbations and Quasi-Representable Relations}

@@

We must now impose an additional condition on predomain relations $c$ and error
domain relations $d$ that specify how they interact with these perturbations. In
particular, we need to be able to ``push'' and ``pull'' perturbations along
relations $c$ and $d$. The intuition for this requirement comes from the
construction of the square corresponding to the original UpL rule from the
square for the simplified version as was shown in Section
\ref{sec:towards-relational-model}. Recall that we horizontally composed the
simplified UpL for $c$ square with the identity square for $c'$ on the right.
Now that the square for UpL involves a delay, this construction doesn't work
unless we can ``push'' the delay on $A_2$ to a delay on $A_3$. That is, given a
relation $c : A \rel A'$ and a perturbation $m_A \in M_A$, we need to be able to
turn it into a perturbation $m_{A'} \in M_{A'}$ such that the resulting
endomorphisms obtained by the respective homomorphisms $i_A$ and $i_{A'}$ form a
square. This is made formal by the following definition:
%
\begin{definition}
    Let $(A, M_A, i_A)$ and $(A', M_{A'}, i_{A'})$ be value objects, and let
    $c : A \rel A'$ be a predomain relation. A \emph{push-pull structure} for
    $c$ consists of a homomorphism $\push : M_A \to M_A'$ and $\pull : M_A' \to
    M_A$ such that for all $m_A \in M_A$ there is a square $i_A(m_A) \ltsq{c}{c}
    i_{A'}(\push\, m_A)$ and for all $m_{A'} \in M_{A'}$ there is a square
    $i_A(\pull\, m_{A'}) \ltsq{c}{c} i_{A'}(m_{A'})$.
\end{definition}
%
We make the analogous definition for computation objects.

With the notion of perturbation defined, we can now discuss the final aspect of
the model, namely the representability of relations.
%
\begin{definition}[quasi-left-representable relations]
Let $(A, M_A, i_A)$ and $(A', M_{A'}, i_{A'})$ be value objects, and let $c : A
\rel A'$ be a predomain relation. We say that $c$ is
\emph{quasi-left-representable} by a predomain morphism $e : A \to A'$ if there
are perturbations $\delle_c \in M_A$ and $\delre_c \in M_{A'}$ such that the
following two squares exist: (1) UpL: $e \ltsq{c}{r(A')} i_{A'}(\delre_c)$, and
(2) UpR: $i_A(\delle_c) \ltsq{r(A)}{c} e$.
\end{definition}
%
Observe that this generalizes the previous notion of representability, since
under that definition the perturbations were always the identity.
%
We make the analogous definition of quasi-left-representability for error domain
relations $d$.
%
Likewise, we define quasi-right-representability as follows:
%
\begin{definition}
Let $(B, M_B, i_B)$ and $(B', M_{B'}, i_{B'})$ be computation objects, and let
$d : B \rel B'$ be an error domain relation. We say that $d$ is
\emph{quasi-right-representable} by an error domain morphism $p : B' \to B$ if
there are perturbations $\dellp_d \in M_B$ and $\delrp_d \in M_{B'}$ such that
the following two squares exist: 
(1) DnL: $p \ltsq{r(B')}{d} i_{B'}(\delrp_d)$ and
(2) DnR: $i_B(\dellp_d) \ltsq{d}{r(B)} p$.
\end{definition}
%
We make the analogous definition of quasi-right-representability for predomain
relations $c$.

@@

We can now give the final definition of value and computation relations for our
model:
%
\begin{definition}[value relations]
A \textbf{value relation} between value objects $(A, M_A, i_A)$ and $(A',
M_{A'}, i_{A'})$ consists of a predomain relation $c : A \rel A'$ that has a
push-pull structure, is quasi-left-representable by a morphism $e_c : A \to A'$,
and is such that $Fc$ is quasi-right-representable by an error domain morphism
$p_c : FA' \to FA$.
\end{definition}

\begin{definition}[computation relations]
A \textbf{computation relation} between computation objects $(B, M_B, i_B)$ and
$(B', M_{B'}, i_{B'})$ consists of an error domain relation $d$ that has a
push-pull structure, is quasi-right-representable by a morphism $p_d : B' \to
B$, and is such that $Ud$ is quasi-left-representable by a morphism $e_d : UB
\to UB'$
\end{definition}

The reason for requiring $Fc$ to be quasi-right-representable in the definition
of value relations is so that we can define the action of the functor $F$ on
relations. Similarly, we require $Ud$ to be quasi-left-representable in the
definition of computation relations so that we can define the action of $U$ on
relations.

It remains to specify the actions of the functors on value and computation
objects and relations.
% Because we have augmented the definitions of object and relation for our model,
% we need to specify the actions of the functors $F$, $U$, $\times$, and $\arr$ on
% these objects and relations. 
For instance, the functor $F$ now acts on
\emph{value objects} $(A, M_A, i_A)$, which means we need to specify the monoid
$M_{FA}$ and interpretation $i_{FA}$ corresponding to the error domain $FA$.

\subsubsection{Actions on Objects}
We first discuss the action of $\li$ on objects. Given a value object $(A, M_A,
i_A)$ we define $M_{\li A}$ to be $\mathbb{N} \oplus M_A$, where $\oplus$
denotes the free product of monoids (the coproduct in the category of monoids).
The intuition behind needing to include $\mathbb{N}$ comes from the example of
the downcast for $\iarr$. Recall that we adjusted the squares corresponding to
the DnL and DnR rules by adding $\delta = \theta \circ \nxt$ on the side
opposite the downcast. Thus, we need a syntactic perturbation in $M_{\li A}$
that will be interpreted as $\delta$. We accomplish this by taking the coproduct
with $\mathbb{N}$.
%
Then to interpret the perturbations as endomorphisms, by the universal property
of the coproduct of monoids it suffices to define two homomorphisms, one from
$\mathbb{N} \to \morbisimid{\li A}$ and one from $M_A \to \morbisimid{\li A}$.
The first homomorphism sends the generator $1 \in \mathbb{N}$ to the endomorpism
$(\delta \circ \eta)^\dagger$, and the second sends $m_A \in M_A$ to
$\li(i_A(m_A))$ and we observe in both cases that the resulting endomorphisms
are bisimilar to the identity since the action of $\li$ on morphisms preserves
bisimilarity.

Now we discuss the action of $U$ on objects. Given a computation object $(B,
M_B, i_B)$, we define $M_{UB}$ to be $\mathbb{N} \oplus M_B$. The reason for
requiring $\mathbb{N}$ is related to the Kleisli arrow functor: to establish
quasi-representability of $U(c \arr d)$ given quasi-representability of $c$ and
$d$, we will need to turn a perturbation on $FA$ into a perturbation on $U(A
\arr B)$. Since the perturbations for $FA$ involve $\mathbb{N}$, so must the
perturbations for $UB$. The interpretation $i_{UB}$ of the perturbations on $UB$
works in the same manner as that of $FA$.

The action of $\times$ on objects is as follows. Given $(A_1, M_{A_1}, i_{A_1})$
and $(A_2, M_{A_2}, i_{A_2})$, we define the monoid $M_{A_1 \times A_2} =
M_{A_1} \oplus M_{A_2}$. To construct the homomorphism $i_{A_1 \times A_2}$, we
appeal to the universal property and give homomorphisms $M_{A_1} \to
\morbisimid{A_1 \times A_2}$ and $M_{A_2} \to \morbisimid{A_1 \times A_2}$.

The action of $\arr$ on objects is as follows. Given $(A, M_A, iA)$ and $(B,
M_B, i_B)$ we define the monoid $M_{A \arr B} = M_A^{op} \oplus M_B$. Then to
construct the interpretation as endomorphisms it suffices to define a
homomorphism $M_A^{op} \to \morbisimid{A \arr B}$ and $M_B \to \morbisimid{A
\arr B}$. The former takes an element $m_A$ and a predomain morphism $f : A \to
UB$ and returns the composition $f \circ i_A(m_A)$, while the latter is defined
similarly by post-composition with $i_B$.

\subsubsection{Actions on Relations}

\subsection{The Dynamic Type}
Lastly, we describe the denotation of the dynamic type. The predomain
representing the dynamic type is defined using a combination of least fixpoint
and guarded recursion as the solution to the equation
% \footnote{In this section, we write $F$ instead of $\li$ so that the notation follows that of the abstract model section.}
%
\[ D \cong \mathbb{N}\, + (D \times D)\, + \laterhs U(D \arr FD). \]
%
% Note that the operators in the above equation are all combinators for predomains, so
% this also defines the ordering and the bisimilarity relation for $D$.

@@

Now we define the perturbations for $D$. Recall that for each value type $A$ we
associate a monoid $M_A$ of perturbations and a homomorphism into the monoid of
endomorphisms bisimilar to the identity, and likewise for computation types. We
define the perturbations for $D$ via least-fixpoint in the category of monoids
as
%
\( M_D \cong (M_{D \times D}) \times M_{U(D \to FD)}. \)
%
Unfolding the definitions, this is the same as
%
\( M_D \cong (M_D \times M_D) \times (\mathbb{N} \times M_D^{op} \times \mathbb{N} \times M_D). \)
%
We now explain how to interpret these perturbations as endomorphisms.
We define $i_D : M_D \to \{ f : D \to D \mid f \bisim \id \}$ below,
% via the universal property of the coproduct of monoids, giving a case for each of the generators.
% In the below, note the use of the functorial action of $\arr$ on morphisms.
%
% \[ P_D \cong (P_D \times P_D) \times (P_D \times (\mathbb{N} \times P_D)), \]
%
\begin{align*}
 i_D(p_{\text{times}}, p_{\text{fun}}) &= \lambda d.\text{case $d$ of}  \\
 &\alt \tnat(m) \mapsto \tnat(m) \\
    &\alt \ttimes(d_1, d_2) \mapsto {\ttimes(i_{D \times D}(p_\text{times})(d_1, d_2))} \\
    &\alt \tfun(\tilde{f}) \mapsto {\tfun(\lambda t. i_{U(D \to FD)}(\tilde{f}_t))}
\end{align*}
    % \item $\ptb_D(1)$
    % \item $\ptb_D(\delta^K_D)$ is defined similarly to the previous but has
    % \[ \id \arr i^K(\delta^K_D) \] instead.
    % \item $\ptb_D(\delta^K_D) = \lambda d.\text{case $d$ of}$
    %   \begin{align*} 
    %     &\alt \ttimes(d_1, d_2) \mapsto {\ttimes(i^K(\delta^K_D)(d_1), d_2)} \\
    %     &\alt d' \to d'
    %   \end{align*}
    % \item $\ptb_D(\delta^K_D)$ is defined similarly to the previous but has 
    % \[ (d_1, \ptb_D(\delta^K_D)(d_2)) \] instead.
One can verify that this defines a homomorphism from $M_D \to \{ f : D \to D : f
\bisim \id \}$. We claim that the three relations $\inat$, $\itimes$, and
$\iarr$ satisfy the push-pull property. As an illustrative case, we establish
the push-pull property for the relation $\iarr$. We define $\pull_{\iarr} : M_D
\to P_{U(D \arr FD)}$ by \( \pull_{\iarr}(p_{\text{times}}, p_{\text{fun}}) =
p_{\text{fun}}, \)
% (recall that $P_{U(D \arr FD)} = \mathbb{N} \times P_D^{op} \times \mathbb{N} \times P_D$).
%
i.e., we simply forget the other perturbation.
%
We define $\push_{\iarr} : M_{U(D \arr FD)} \to M_D$ by \( \push_{\iarr}(p_{\text{fun}}) = (\id, p_{\text{fun}}) \).
%
Showing that the relevant squares commute is straightforward.
%
% To do so, consider an arbitrary perturbation $p$ on $D$. 
% Let $(f, d) \in \iarr$. This means that $d$ must be of the form $\tfun{\tilde{f}}$. 
% When we interpret the perturbation on $D$, obtaining and endomorphism to which we then apply $d$,
% we will be perform the action that was performed by the perturbation on the other side and thus
% we will be done by our assumption that $f$ is related to $d$.

@@

\subsection{Defining an Extensional Model}

Lastly, we apply the ``bisimilarity-closure'' construction introduced in the
previous section as follows: In particular, let $c_i : A_i \rel A_i'$ and $c_o :
A_o \rel A_o'$.
%
% We define an extensional square between morphisms $f : A_i \to A_o$ and $g :
% A_i' \to A_o'$ to consist of:

% \begin{itemize}
%   \item A morphism $f' : A_i \to A_o$ with $f \bisim f'$.
%   \item A morphism $g' : A_i' \to A_o'$ with $g \bisim g'$.
%   \item A square $f' \ltsq{c_i}{c_o} g'$.
% \end{itemize}
%
We say that $f \ltsqbisim{c^i}{c_o} g$ if there
exist $f'$ and $g'$ such that
%
\[ f \bisim f' \ltsq{c_i}{c_o} g' \bisim g. \]
%
We make the analogous construction for the computation squares. If the relations
$c_i$ and $c_o$ are obvious from the context, we will simply write $f \ltbisim
g$.

% Now that we have defined an intensional model with an interpretation for the
% dynamic type, we can apply the abstract constructions introduced in Section
% \ref{sec:extensional-model-construction}. Doing so, we obtain an extensional
% model of gradual typing, where the squares are given by the ``bisimilarity
% closure'' of the intensional error ordering.

@@

In this section, we prove an adequacy result for the extensional model of
gradual typing we obtained in the previous section. First we establish some
notation. Fix a morphism $f : 1 \to \li \mathbb{N} \cong \li \mathbb{N}$. We
write $f \da n$ to mean that there exists $m$ such that $f = \delta^m(\eta n)$
and $f \da \mho$ to mean that there exists $m$ such that $f = \delta^m(\mho)$.

Recall that $\ltbisim$ denotes the relation on value morphisms defined as the
bisimilarity-closure of the intensional error-ordering on morphisms. That is, we
have $f \ltbisim g$ iff there exist $f'$ and $g'$ with
%
\[ f \bisim_{\li \mathbb{N}} f' \le_{\li \mathbb{N}} g' \bisim_{\li \mathbb{N}} g. \]
%
Here $\le_{\li \mathbb{N}}$ is the lock-step error ordering, and $\bisim_{\li
\mathbb{N}}$ is weak bisimilarity. First observe that in this ordering, the
semantics of error is not equivalent to the semantics of the diverging term. The
main result we would like to show is as follows: If $f \ltbisim g : \li \mathbb{N}$, then:
\begin{itemize}
  \item If $f \da n$ then $g \da n$.
  \item If $g \da \mho$ then $f \da \mho$.
  \item If $g \da n$ then $f \da n$ or $f \da \mho$.
\end{itemize}

%
Unfortunately, this is actually not provable!
Roughly speaking, the issue is that this is a ``global'' result, and it is not possible
to prove such results inside of the guarded setting. 
In particular, if we tried to prove the above result in the guarded
setting, we would run into a problem where we would have a natural number
``stuck'' under a $\later$, with no way to get out the underlying number.
%
Thus, to prove our adequacy result, we need to leave the guarded setting and pass back
to the more familiar, set-theoretic world with no internal notion of step-indexing.
We can do this using a process known as \emph{clock quantification}.
Recall that all of the constructions we have made in SGDT take place in the context of a clock $k$.
All of our uses of the later modality and guarded recursion have taken place with respect to this clock.
For example, recall the definition of the lift monad by guarded recursion.
% We define the lift monad $\li^k X$ as the guarded fixpoint of $\lambda \tilde{T}. X + 1 + \later^k_t (\tilde{T}_t)$.
We can view this definition as being parameterized by a clock $k$: $\li^k : \type \to \type$.
Then for $X$ satisfying a certain technical requirement known as \emph{clock-irrelevance},
\footnote{A type $X$ is clock-irrelevant if there is an isomorphism $\forall k.X \cong X$.}
we can define the ``global lift'' monad as $\li^{gl} X := \forall k. \li^k X$.

### 2024-07-09 paper-new/concrete-relational-model.tex
work on relational model

\subsubsection{Modelling Type Precision}
We model type precision $c : A \ltdyn A'$ as a \emph{monotone relation}, i.e., a
relation $c$ that is upward-closed under the relation on $A'$ and
downward-closed under the relation on $A$. We denote such a relation between $A$
and $A'$ by $c : A \rel A'$. The relation on the poset $A$ is denoted $r(A)$.
% This extends to products $c_1 \times c_2$ in the obvious way.
Composition of relations on predomains is the usual relational composition
(which is truncated to be propositional).

>>>>

We model a type precision relation $c : A \ltdyn A'$ as a \emph{monotone relation}, i.e., a
relation $c$ that is upward-closed under the relation on $A'$ and
downward-closed under the relation on $A$. We denote such a relation between $A$
and $A'$ by $c : A \rel A'$. The relation on the poset $A$ is denoted $r(A)$.
% This extends to products $c_1 \times c_2$ in the obvious way.
Composition of relations on predomains is the usual relational composition
(which is truncated to be propositional).
